### 2021-08-19 reviewer_answers/Response to reviewers.tex

\item
The diffusion (second spatial derivative) term in the left-hand side of equation (2.2) represents the evolutionary process. I would call g the diffusion tensor instead of the diffusion coefficient (line 180).

AUTHOR answer: This has been corrected. We searched for other instances of "diffusion coefficient" aswell, but there were none.

\item
Since g is the diffusion tensor. It is assumed to be symmetric (Assumption 3.1). Of course, this symmetry facilitates the mathematical treatment with respect to well-posedness and error estimation. It is said that in the manuscript that Assumptions 3.1 follow from biology. Could there be any biological application cases where g is nonsymmetric? \\

AUTHOR answer: 
In thermodynamical systems, whenever sub-scale stochastic processes are modeled as diffusive on a larger scale, one can typically establish that diffusion tensor is symmetric and positive definite due to Onsager's reciprocal relations. To be fair, we are not aware of any analysis establishing the validity of Onsager's reciprocal relations in the current context, and it is conceptually possible that the diffusion tensor might be nonsymmetric. However, we are not aware of any biological system in which this is the case.

A different perspective on symmmetry is that it implies that there exists a coordinate map of the trait space such that diffusion is isotropic. Conceptually, one could imagine this map being onto the genetic space, and argue that mutations of genes are all equiprobable and thus isotropic. If one accepts this perspective, then the symmetry of the diffusion tensor is a consequence of the pullback of the mapping.

In other words, we strongly believe that the symmetry of the diffusion tensor is really symmetric, but in absence of a conclusive analysis, we have updated the language in the manuscript to be less absolute with respect to this point.

\item
Regarding the boundary condition that n vanishes on the boundary of the trait space (assuming that the trait space is 'large enough'), one could also relax this BC by considering g grad n + A n = 0. Although this BC would relax that n = 0, one has to fiddle around with the A-parameter. Do you think that this alternative BC could do a better job?

AUTHOR answer: 
We agree that this is a possible, and most likely the a posteriori analysis would adapt naturally to this approach. However, this introduces a coefficient that most likely is variable along the full boundary of the local domain. As it is anyway a necessary conceptual assumption that a species is relatively local in trait space, we believe that it is conceptually more natural to capture the whole species by introducing a sufficiently large local domain such that zero BCs are satisfactory.

@@

\item
Regarding the modelling part. The g-tensor for diffusion is based on 'ordinary' diffusion in which random changes are gradual. In many biological systems, mutations could give large, sudden changes over time, which could be modelled using Levy-jump processes. Then one has to incorporate fractional derivatives. Perhaps the Authors could comment on this. (I am not asking the Authors to carry out the analysis for this case, that would really be too much, the question is just meant to be philosophical).

AUTHOR answer: Yes, we agree, conceptually fractional derivatives (in place of the gradient operator) can be an attractive modeling option. We have included a comment to this effect.

@@

\item
The multi-species model models the transition between species, and in a short subsection (2.3), the relation between the multi-species model and the polultion level model is presented. What I miss om the multi-species model is the actual evolutionary part of the process. There is no random part in this model, which does exist in the population level model through the volatility (diffusion) part. Should not the Authors include a Wiener process (or Levy process to account for jump processes) here (to allow random walk)? Now the approach seems entirely deterministic.

AUTHOR answer: It is correct that the model considered is fully deterministic. We consider this as a question of modeling scales (with the analogy to particle and Boltzmann models for gas dynamics): Deterministic mean-field theories are suitable at sufficiently large scales, which in this context implies that the population densities sufficiently large that the full stochastic process is sampled. Conversely, if small populations are considered (such as rare species, or similar), the particular instances of the stochastic process will be essential, and stochastic equations are a natural modeling framework.

@@

\item
Line 528 please, replace 'The figure 2 below' with 'Figure 2 below'

AUTHOR answer: This has been corrected.

\item
Your analysis is also built on the existence of statistical moments, hence in the case of a Cauchy distribution (which has a very 'normal' appearance), the model may lead to problems. It could be interesting to do a numerical experiment with a Cauchy distribution function. ($f(x) = 1/\pi \cdot 1 / (x^2 + 1)$ is the standard Cauchy distribution, which has no moments (average, standard deviation ...).

AUTHOR answer: The reviewer raises a subtle point which essentially asks how traits should be measured at the species level.

The derivation of the species level model from the population level model rests on two fundamental assumptions; (1) that distinct species can be identified at all times over the global population distribution (i.e., as compactly supported, non-overlapping sub-distributions), and (2) that these sub-distributions are of a known statistical quality (in this case, the Gaussian/normal distribution). As discussed in [1] (where the species level model is derived), the choice of the Gaussian was to a certain degree due to convenience (although individual traits are commonly assumed to follow a normal distribution in the biological literature), hence other statistical distributions could also be accommodated (at the cost of an increasingly complex calculation/notation, see e.g., page 5-6 in [1] where the possibility of using the uniform distribution is briefly discussed). However, for the exact reasons given in your comment, the Cauchy distribution is not a possible choice, although Cauchy distributions do not pose a problem in terms of the population level model. \\

However, the question about the shape of the distribution is somewhat misleading: The a posteriori error estimates are intended to assess the validity of the (reduced) species level model, be it good or bad. It becomes clear, by applying any monotone transformation of the axis of the trait space (such as e.g., the logarithm), that in the process of setting up the model, one makes a choice as to how to measure traits. Most classically, one can measure the trait "weight" on a linear or logarithmic scale, and even on a cube-root scale (via the proxy variable "length"). Clearly, the trait distribution will not be Gaussian for more than at most one of these choices. With a poor choice of scale for a trait, where e.g., the trait follows a Cauchy distribution, the species level model will in general be poor, and conceptually the a posteriori error estimates will indicate that the population-level model should always be applied, and no multi-scale algorithm is possible.

To a certain extent, this is completely analogous to what we see during the speciation event. At this point, the distribution becomes far from Gaussian, and the error estimates indicate that the species level model is not suitable.

In practice, this implies that for any real modeling of a real physical system, one should advice the biologist to choose a measurement scale for which traits are observed to be close to normally distributed, and provide a warning that if this advice is not heeded, then disappointing results can come from the subsequent analysis.

We have added comments to this effect in the fifth paragraph of the introduction (pages 2-3), where we discuss the species and population level models, we have added a few sentences outlining the above point concerning assumption (2). We have also added in section (2.3) that here $n$ should be smooth enough for these moments to exist.\\

@@

AUTHOR answer: Thank you for the positive evaluation of our manuscrupt, we really appreciated your thoughtful comments!
\end{enumerate}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\textbf{Referee \#2  (Remarks to the Author):}\\

This manuscript is concerned with a mathematical modelling for speciation using a general Lotka-Volterra equation modelling the evolution of some ecosystem. This model takes into account species traits and mutations through a diffusion term. This general equation is coupled with a system of ODE modelling the species level: density of species together with the evolution of the mean traits and the covariance matrix.

\begin{enumerate}
\item
The authors discuss and propose a model for the speciation processes leading to new species. To do so, they make use of a discretized version of the above equations. The discussions on the speciation processes seems to be highly related to the choice of the discrete system and in particular on the reconstruction operator. No connection with the original continuous system is clearly explained neither how it depends on the discretization.
Speciation is modeled in term of clustering in the trait space. Does it depend on the traits kept in the model?

AUTHOR answer: Indeed, there was missing a discussion on in what sense the time-discrete system relates to the original continuous one. To rectify this, we have rewritten the first paragraph of Section 3 (p. 7) where we elaborate on this. Hopefully, this relation is now clear, and the impression that the results are highly related to the time-discrete formulation (which is simply a standard backward Euler approximation to the continuous system), has been rectified.

As to the clustering in trait space: Yes this depends not just on the traits chosen to model the system, but also on their scale of measurement. We have added some discussion to this effect in the introduction relating also to a comment by the first reviewer.

### 2021-08-19 manuscript/ex_article.tex
last round of changes

The species-interaction model we take as the starting point is formulated as a system of ordinary differential equations (ODE) describing the dynamics of a biological system in which any finite number of species interact and evolve adaptively over evolutionary time (referred to in the sequel as the \emph{species level}, SLM, or \emph{macro-scale} model). This model extends the Lotka-Volterra system by representing each species not only by its abundance, but also by its mean traits coordinate and trait covariance matrix. It originates in a recent work by two of the authors~\cite{nordbotten2016asymmetric}, wherein the model equations were derived by applying a moment closure averaging technique on a deterministic trait-specific population density model (referred to in the sequel as the \emph{population level}, PLM, or \emph{micro-scale} model), \red{under the assumptions that (A1) distinct species can be identified at all times, and (A2) that the species distributions are of a known statistical quality (in this case, the normal distribution)}. See also related derivations in e.g.,~\cite{debarre2014multidimensional,sasaki2011oligomorphic}. The population level model is formulated as a partial differential equation (PDE), and governs the abundance density as a function of evolutionary traits. Moreover, the authors showed in~\cite{nordbotten2016asymmetric} that the species and population level models are consistent, i.e., they describe the same system dynamics as long as the system behavior is such that both models are valid. Still, these descriptions are fundamentally different in that they operate on separate scales; the species level model operates with the coarser unit of distinct species, while the population level model operates with the finer unit of the individual (i.e., the category of species is not imposed upon the individuals who make up the population). Thus, the species level model described above approximates the system dynamics governed by the population level model, under the conditions that the population can be grouped into distinct species, \red{and that each species' distribution conforms to a normal distribution}. Note also that due to the deterministic nature of the population level model, this property is also inherited at the species level.

@@

Given an open, bounded domain $\Omega \subset \real^d$ in trait space (i.e., the number of evolutionary traits considered is equal to $d \geq 1$), and final time $T>0$, let $\Omega_T := \Omega \times (0,T)$ be the space-time domain, and let $n: \Omega_T \rightarrow \real$ be the trait-specific population density, or, number of individuals per measure on $\real^d$. The domain $\Omega$ can then be regarded as all evolutionary traits attainable by the relevant population over evolutionary time, and the total number of individuals present at any time $t \in (0,T)$ is then given by integration over all attainable traits, i.e., $\int_\Omega n(x,t) \dx$, where we require non-negativity of the population density, i.e., 
\begin{equation}
n \geq 0, \textn{ for all } (x,t) \in \Omega_T.
\end{equation}
Furthermore, let $f: \Omega_T \rightarrow \real$ be the source term (net rate of migration/immigration). The model then reads as the following nonlinear and non-local equation:
\begin{equation}
\pt n - rn + \Phi(n)n -\nabla \cdot(g\nabla n) = f. \label{PDE} 
\end{equation}
The ecological processes consists of the growth term, $-rn$, where $r : \Omega_T \rightarrow \real$ is the inherent per-capita growth-rate, and the interaction term, 
\begin{equation}\label{def_interaction}
\Phi(n)n := bn^2 - n \int_\Omega \alpha(x,y,t)n(y,t)\dy,
\end{equation}
where $b : \Omega_T \rightarrow \real_{\geq 0}$ is the local individual limitation coefficient, 
and $\alpha : \Omega \times \Omega_T \rightarrow \real$ is the non-local individual interaction coefficient, i.e., $\alpha(x,y,t)$ is the effect of an individual with trait coordinate $y$ on an individual with trait coordinate $x$. The diffusive term represents the evolutionary process, where $g : \Omega_T \rightarrow \real^{d\times d}$ is the intergenerational trait diffusion \red{tensor}. Generally speaking, the model equation \eqref{PDE} belongs to the class of reaction-diffusion type parabolic conservation equations (in this case, with a non-local reaction term). Since homogenous boundary conditions of Dirichlet type is always justified from a biological perspective given a large enough trait domain $\Omega$, we shall only consider this situation for the present purposes. We let initial data be specified by $n(x,0) = n_0(x)$. The PLM then reads as the following initial/boundary value problem:

@@

Find $N :(0,T) \rightarrow \real^{s(1+d+d^2)}$ such that $n_i \geq 0$, for $1\leq i \leq s$, and 
\bse
\begin{alignat}{2}
\dfrac{\textnormal{d}N}{\textnormal{d}t} &= \Ncal(N), \quad && t\in(0,T), \label{ODEmodel}\\
N &= N_0, &&t = 0. \label{ODEmodelIC}
\end{alignat}
\ese
In practice, the problem \eqref{ODEmodel}--\eqref{ODEmodelIC} can be reduced to $s(1+d+d(d+1)/2)$ equations due to the symmetry of the trait covariance matrix. For additional details including a derivation of the above model, see~\cite{nordbotten2020dynamics}. Note also that when indexing the components of a point $x\in\Omega$ we use subscripts, i.e., $x = (x_1,\cdots,x_d)$, not to be confused with the species mean trait coordinate $x_i$, where we write $x_i(t) = (x_{i,1}(t),\cdots,x_{i,d}(t))$. 
%The distinction should be clear from the context.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Relationship between population and species level models}
If there is a time interval $J := (t_{0},t_1) \subset (0,T)$ where $s \geq 1$ species can unambiguously be identified, there exists a collection of non-overlapping (possibly time dependent), open and bounded regions in trait space, i.e., $\{B_i \subset \Omega, i \leq s,  B_i \cap B_j = \emptyset, i \neq j\}$, where the population density is compactly supported and contains only one local maximum. Thus, if we let $\mu_{B_i}^k(n)$, be the $k$'th order moment of $n$ (for $k \in \{0,1,2\}$, \red{and assuming $n$ is smooth enough for these moments to exist}), within the region $B_i$, i.e.,
\bse
\begin{alignat}{1}
\mu_{B_i}^0(n) &:= \int_{B_i} n(x)\dx,\label{m0}\\ 
\mu_{B_i}^1(n) &:= \frac{1}{\mu_{B_i}^0(n)} \int_{B_i} n(x)x\dx,\label{m1}\\
\mu_{B_i}^2(n) &:= \frac{1}{\mu_{B_i}^0(n)} \int_{B_i} n(x)(x-\mu_{B_i}^1)\otimes (x-\mu_{B_i}^1)\dx,\label{m2}
\end{alignat}
\ese
we can initialize the SLM at $t=t_{0}$ with initial data
\begin{equation}
N_{0} = (\mu_{B_{1}}^0,\mu_{B_{1}}^1,\mu_{B_{1}}^2,\cdots,\mu_{B_{s}}^0,\mu_{B_{s}}^1,\mu_{B_{s}}^2)|_{t=t_0}.
\end{equation}
Solving \eqref{ODEmodel}--\eqref{ODEmodelIC} then amounts to approximating \eqref{m0}--\eqref{m2} for $t \in J$.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Time discrete PLM}\label{sec:timedisc}
In this section we introduce the time discrete PLM, which will serve as the basis for the a posteriori error bounds. First, we establish the existence of a solution for every discrete time, then we derive an identity for the residual which also provides uniqueness of the time-discrete solution.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Notation}
For $k = 0,\cdots,M$, we let $t_k$ denote the discrete times, such that $0 = t_0 < t_1 < \cdots < t_M := T$, and where $t_k = t_{k-1} + \tau_k = \sum_{j = 1}^k \tau_k$, for some set of time increments $\{\tau_k\}_{k=1}^M$. In the following, we use superscripts to indicate dependency on the discrete times, e.g., $u^k := u(t_k)$.

We employ standard notation for function spaces. For a function space $V$ we denote by $\norm{\cdot}_V$ its energy norm, by $V^\ast$ its dual space, and by $\langle \cdot, \cdot \rangle_{V^\ast,V}$ the dual pairing. In particular, for $1 \leq p \leq \infty$, and a domain $D \subset \Omega$ we denote by $L^p(D)$ the Lebesgue-spaces of integrable functions defined on $D$, with associated norm $\norm{\cdot}_{p,D}$ (domain subscript omitted if this is clear from the context). Furthermore, let $L^p_{+}(D) := \{u\in L^p(D) : u \geq 0 \textn{ a.e.}\}$. For the case $p=2$, we let $\norm{u} := \norm{u}_{2} = (u,u)^{\frac{1}{2}}$, where $(\cdot,\cdot)$ is the standard $L^2$ inner product. Moreover, we denote by $\Hdiv$ the space of functions in $[L^2(D)]^d$ admitting a weak divergence, and by $H^1(D)$ the space of functions in $L^2(D)$ admitting weak gradients, with its vanishing trace subspace denoted by $H_0^1(D)$, where $H^{-1}(D) := H_0^1(D)^\ast$. Finally, denote by $H_{0,+}^1(D) := H_0^1(D) \cap L^2_{+}(D)$.

@@

Assumption \ref{assum:coeff} is natural from a biological point of view. On the other hand, Assumption \ref{assum:analysis} is introduced for the analysis, and may be more restrictive than desired for certain biological systems. We will return to this issue in the numerical examples (cf. Section~\ref{sec:ex}).

@@

\begin{figure}[h]
  \centering
\begin{subfigure}{.45\textwidth}
  \centering
  \includegraphics[width=1.\linewidth]{figures/ex2/pred_res}
  \caption{Multi-scale method.}
\end{subfigure}%
\hspace{.05\textwidth}
\begin{subfigure}{.45\textwidth}
  \centering
  \includegraphics[width=1.\linewidth]{figures/ex2/pred_res_H}
  \caption{Heuristic method.}
\end{subfigure}  
  \caption{A posteriori modeling-remainder estimator, $\eta_{\textn{rem},i}^{k}$, for `predator'-species as a function of time.}
  \label{ex2:pred_residuals}
\end{figure}
\red{
\begin{remark}[Computation times]
The heuristic method will have a computation time approximately equal to that of solving only the species level model for the same time interval. On the other hand, the multi-scale method will have a computation time higher than that of solving the population level model for a time interval equal to the speciation interval (with the same grid size as the local density region) since the compression operator is implemented as an iteration procedure. Thus, the heuristic method will in general be orders of magnitude faster than the multi-scale method. For this reason, we omit listing a comparison.
\end{remark}
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}
\label{sec:conclusions}
We have developed two strategies for modeling speciation \\ events within the context of species interaction models. The first, \emph{heuristic approach}, is based on splitting the diverging species according to the spectrum of the trait covariance matrix. The second, \emph{multi-scale approach}, is based on resolving the diverging species as a population density distribution using a fine-scale population level model for the duration of the speciation event (i.e., until `child'-species are sufficiently separated in trait space). Crucial to both these approaches is the connection between the species scale and the population scale, i.e., the ability to view the species either as an abundance-trait-covariance tuple, or as a population density distribution. This allows for defining a multi-scale framework in which these two scales are coupled, and to calculate the a posteriori error bound of the reconstructed macro-scale solution which then indicates the modeling error. We have also given conditions on the nonlinearities of the micro-scale model for which well-posedness of the time-discrete problem is guaranteed. Using explicit equilibrated flux and density reconstructions, we presented a posteriori error estimates for an error measure composed of an energy $H^1$-norm and a semi-metric in terms of a  residual monotone operator. In particular, the dual norm of the residual is found to be equal to the error between the exact and approximate solutions, again given conditions on the nonlinearities. Even when the theoretical conditions are not fulfilled, our framework provides a working algorithm in practice, as our second numerical example shows. Finally, regarding the heuristic and multi-scale methods, by comparison of the species parameters with the corresponding moments from the reference (global) PLM solution, it is clear that the multi-scale approach is superior to the heuristic approach. In fact, from our experiments it appears that a multi-scale approach to speciation is indeed required for eco-evolutionary modeling at the species level, in which speciation events are allowed. We propose that this multi-scale approach might serve as a productive way of integrating ecological processes and evolutionary processes.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

The species-interaction model we take as the starting point is formulated as a system of ordinary differential equations (ODE) describing the dynamics of a biological system in which any finite number of species interact and evolve adaptively over evolutionary time (referred to in the sequel as the \emph{species level}, SLM, or \emph{macro-scale} model). This model extends the Lotka-Volterra system by representing each species not only by its abundance, but also by its mean traits coordinate and trait covariance matrix. It originates in a recent work by two of the authors~\cite{nordbotten2016asymmetric}, wherein the model equations were derived by applying a moment closure averaging technique on a deterministic trait-specific population density model (referred to in the sequel as the \emph{population level}, PLM, or \emph{micro-scale} model), \red{under the assumptions that (A1) distinct species can be identified at all times, and (A2) that the species distributions are of a known statistical quality (in this case, the normal distribution)}. See also related derivations in e.g.,~\cite{debarre2014multidimensional,sasaki2011oligomorphic}.

The population level model is formulated as a partial differential equation (PDE), and governs the abundance density as a function of evolutionary traits. Moreover, the authors showed in~\cite{nordbotten2016asymmetric} that the species and population level models are consistent, i.e., they describe the same system dynamics as long as the system behavior is such that both models are valid. Still, these descriptions are fundamentally different in that they operate on separate scales; the species level model operates with the coarser unit of distinct species, while the population level model operates with the finer unit of the individual (i.e., the category of species is not imposed upon the individuals who make up the population). Thus, the species level model described above approximates the system dynamics governed by the population level model, under the conditions that the population can be grouped into distinct species, \red{and that each species' distribution conforms to a normal distribution. While this requirement that trait distributions of a species are essentially normally distributed may seem restrictive, we note that the shape of a distribution is not an intrinsic quality of the observed system, but is equally dependent on the scale of measurement. This assumption can thus be considered not as much as a limitation on the biological system, as a constraint on the scale of measurement.} Note also that due to the deterministic nature of the population level model, this property is also inherited at the species level.

@@

Given an open, bounded domain $\Omega \subset \real^d$ in trait space (i.e., the number of evolutionary traits considered is equal to $d \geq 1$), and final time $T>0$, let $\Omega_T := \Omega \times (0,T)$ be the space-time domain, and let $n: \Omega_T \rightarrow \real$ be the trait-specific population density, or, number of individuals per measure on $\real^d$. The domain $\Omega$ can then be regarded as all evolutionary traits attainable by the relevant population over evolutionary time, and the total number of individuals present at any time $t \in (0,T)$ is then given by integration over all attainable traits, i.e., $\int_\Omega n(x,t) \dx$, where we require non-negativity of the population density, i.e., 
\begin{equation}
n \geq 0, \textn{ for all } (x,t) \in \Omega_T.
\end{equation}
Furthermore, let $f: \Omega_T \rightarrow \real$ be the source term (net rate of migration/immigration). The model then reads as the following nonlinear and non-local equation:
\begin{equation}
\pt n - rn + \Phi(n)n -\nabla \cdot(g\nabla n) = f. \label{PDE} 
\end{equation}
The ecological processes consists of the growth term, $-rn$, where $r : \Omega_T \rightarrow \real$ is the inherent per-capita growth-rate, and the interaction term, 
\begin{equation}\label{def_interaction}
\Phi(n)n := bn^2 - n \int_\Omega \alpha(x,y,t)n(y,t)\dy,
\end{equation}
where $b : \Omega_T \rightarrow \real_{\geq 0}$ is the local individual limitation coefficient, 
and $\alpha : \Omega \times \Omega_T \rightarrow \real$ is the non-local individual interaction coefficient, i.e., $\alpha(x,y,t)$ is the effect of an individual with trait coordinate $y$ on an individual with trait coordinate $x$. The diffusive term represents the evolutionary process, where $g : \Omega_T \rightarrow \real^{d\times d}$ is the intergenerational trait diffusion \red{tensor}. Generally speaking, the model equation \eqref{PDE} belongs to the class of reaction-diffusion type parabolic conservation equations (in this case, with a non-local reaction term). \red{We remark that several generalizations of \eqref{PDE} are possible.  In particular, the diffusive term as stated models incremental evolutionary processes: To allow for rare but possibly non-incremental evolutionary processes more general non-local operators (such as e.g., fractional derivatives), should be considered.}

Since homogeneous boundary conditions of Dirichlet type is always justified from a biological perspective given a large enough trait domain $\Omega$, we shall only consider this situation for the present purposes. We let initial data be specified by $n(x,0) = n_0(x)$. The PLM then reads as the following initial/boundary value problem:

@@

Find $N :(0,T) \rightarrow \real^{s(1+d+d^2)}$ such that $n_i \geq 0$, for $1\leq i \leq s$, and 
\bse
\begin{alignat}{2}
\dfrac{\textnormal{d}N}{\textnormal{d}t} &= \Ncal(N), \quad && t\in(0,T), \label{ODEmodel}\\
N &= N_0, &&t = 0. \label{ODEmodelIC}
\end{alignat}
\ese
In practice, the problem \eqref{ODEmodel}--\eqref{ODEmodelIC} can be reduced to $s(1+d+d(d+1)/2)$ equations due to the symmetry of the trait covariance matrix. For additional details including a derivation of the above model, see~\cite{nordbotten2020dynamics}. Note also that when indexing the components of a point $x\in\Omega$ we use subscripts, i.e., $x = (x_1,\cdots,x_d)$, not to be confused with the species mean trait coordinate $x_i$, where we write $x_i(t) = (x_{i,1}(t),\cdots,x_{i,d}(t))$. 
%The distinction should be clear from the context.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Relationship between population and species level models}
If there is a time interval $J := (t_{0},t_1) \subset (0,T)$ where $s \geq 1$ species can unambiguously be identified, there exists a collection of non-overlapping (possibly time dependent), open and bounded regions in trait space, i.e., $\{B_i \subset \Omega, i \leq s,  B_i \cap B_j = \emptyset, i \neq j\}$, where the population density is compactly supported and contains only one local maximum. Thus, if we let $\mu_{B_i}^k(n)$, be the $k$'th order moment of $n$ (for $k \in \{0,1,2\}$, \red{and assuming $n$ is smooth enough for these moments to exist}), within the region $B_i$, i.e.,
\bse
\begin{alignat}{1}
\mu_{B_i}^0(n) &:= \int_{B_i} n(x)\dx,\label{m0}\\ 
\mu_{B_i}^1(n) &:= \frac{1}{\mu_{B_i}^0(n)} \int_{B_i} n(x)x\dx,\label{m1}\\
\mu_{B_i}^2(n) &:= \frac{1}{\mu_{B_i}^0(n)} \int_{B_i} n(x)(x-\mu_{B_i}^1)\otimes (x-\mu_{B_i}^1)\dx,\label{m2}
\end{alignat}
\ese
we can initialize the SLM at $t=t_{0}$ with initial data
\begin{equation}
N_{0} = (\mu_{B_{1}}^0,\mu_{B_{1}}^1,\mu_{B_{1}}^2,\cdots,\mu_{B_{s}}^0,\mu_{B_{s}}^1,\mu_{B_{s}}^2)|_{t=t_0}.
\end{equation}
Solving \eqref{ODEmodel}--\eqref{ODEmodelIC} then amounts to approximating \eqref{m0}--\eqref{m2} for $t \in J$.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Time discrete PLM}\label{sec:timedisc}
In this section we introduce the time discrete PLM, which will serve as the basis for the a posteriori error bounds: 
\red{
For $k = 0,\cdots,M$, we let $t_k$ denote the discrete times, such that $0 = t_0 < t_1 < \cdots < t_M := T$, and where $t_k = t_{k-1} + \tau_k = \sum_{j = 1}^k \tau_k$, for some set of time increments $\{\tau_k\}_{k=1}^M$ (in the following, we use superscripts to indicate dependency on the discrete times, e.g., $u^k := u(t_k)$). The time-discrete version of eq.~\ref{PDE} then reads as (for $k\geq 1$)
\begin{equation}
n^k - r^k n^k + \Phi^k(n^k)n^k - \nabla \cdot (g^k \nabla n^k) = f + n^{k-1},\label{timediscPDE}
\end{equation}
where we have discretized in time using a standard implicit first order method (i.e., the backward Euler method). While we do not prove convergence of the above discrete scheme to the original continuous problem, the local truncation error for the backward Euler method is known to be of second order, and one would expect first-order convergence of the time-discrete problem to the continuous problem for sufficiently smooth problems. }

\red{We remark that the time-discrete problem has significant biological relevance by itself, as many processes are naturally modeled on a discrete time-scale due to the presence of strong temporal cycles (e.g., day, year, and generation).}

\red{In the following, we first establish the existence of a solution to \eqref{timediscPDE} in the weak sense for every discrete time, then we derive an identity for the residual which also provides uniqueness of the time-discrete solution.}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Notation}
We employ standard notation for function spaces. For a function space $V$ we denote by $\norm{\cdot}_V$ its energy norm, by $V^\ast$ its dual space, and by $\langle \cdot, \cdot \rangle_{V^\ast,V}$ the dual pairing. In particular, for $1 \leq p \leq \infty$, and a domain $D \subset \Omega$ we denote by $L^p(D)$ the Lebesgue-spaces of integrable functions defined on $D$, with associated norm $\norm{\cdot}_{p,D}$ (domain subscript omitted if this is clear from the context). Furthermore, let $L^p_{+}(D) := \{u\in L^p(D) : u \geq 0 \textn{ a.e.}\}$. For the case $p=2$, we let $\norm{u} := \norm{u}_{2} = (u,u)^{\frac{1}{2}}$, where $(\cdot,\cdot)$ is the standard $L^2$ inner product. Moreover, we denote by $\Hdiv$ the space of functions in $[L^2(D)]^d$ admitting a weak divergence, and by $H^1(D)$ the space of functions in $L^2(D)$ admitting weak gradients, with its vanishing trace subspace denoted by $H_0^1(D)$, where $H^{-1}(D) := H_0^1(D)^\ast$. Finally, denote by $H_{0,+}^1(D) := H_0^1(D) \cap L^2_{+}(D)$.

@@

Assumption \ref{assum:coeff} is\red{, to the best of the authors' understanding,} natural from a biological point of view. On the other hand, Assumption \ref{assum:analysis} is introduced for the analysis, and may be more restrictive than desired for certain biological systems. We will return to this issue in the numerical examples (cf. Section~\ref{sec:ex}).

@@

\begin{figure}[h]
  \centering
\begin{subfigure}{.45\textwidth}
  \centering
  \includegraphics[width=1.\linewidth]{figures/ex2/pred_res}
  \caption{Multi-scale method.}
\end{subfigure}%
\hspace{.05\textwidth}
\begin{subfigure}{.45\textwidth}
  \centering
  \includegraphics[width=1.\linewidth]{figures/ex2/pred_res_H}
  \caption{Heuristic method.}
\end{subfigure}  
  \caption{A posteriori modeling-remainder estimator, $\eta_{\textn{rem},i}^{k}$, for `predator'-species as a function of time.}
  \label{ex2:pred_residuals}
\end{figure}
\red{
\begin{remark}[Computation times]
The heuristic method will have a computation time approximately equal to that of solving only the species level model for the same time interval. On the other hand, the multi-scale method will have a computation time higher than that of solving the population level model for a time interval equal to the speciation interval (with the same grid size as the local density region) since the compression operator is implemented as an iteration procedure. Thus, the heuristic method will in general be orders of magnitude faster than the multi-scale method. 
\end{remark}
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}
\label{sec:conclusions}
We have developed two strategies for modeling speciation \\ events within the context of species interaction models. The first, \emph{heuristic approach}, is based on splitting the diverging species according to the spectrum of the trait covariance matrix. The second, \emph{multi-scale approach}, is based on resolving the diverging species as a population density distribution using a fine-scale population level model for the duration of the speciation event (i.e., until `child'-species are sufficiently separated in trait space). Crucial to both these approaches is the connection between the species scale and the population scale, i.e., the ability to view the species either as an abundance-trait-covariance tuple, or as a population density distribution. This allows for defining a multi-scale framework in which these two scales are coupled, and to calculate the a posteriori error bound of the reconstructed macro-scale solution which then indicates the modeling error. We have also given conditions on the nonlinearities of the micro-scale model for which well-posedness of the time-discrete problem is guaranteed. Using explicit equilibrated flux and density reconstructions, we presented a posteriori error estimates for an error measure composed of an energy $H^1$-norm and a semi-metric in terms of a  residual monotone operator. In particular, the dual norm of the residual is found to be equal to the error between the exact and approximate solutions, again given conditions on the nonlinearities. Even when the theoretical conditions are not fulfilled, our framework provides a working algorithm in practice, as our second numerical example shows. Finally, regarding the heuristic and multi-scale methods, by comparison of the species parameters with the corresponding moments from the reference (global) PLM solution, it is clear that the multi-scale approach is superior to the heuristic approach. In fact, from our experiments it appears that a multi-scale approach to speciation is indeed required for eco-evolutionary modeling at the species level, in which speciation events are allowed. We propose that this multi-scale approach might serve as a productive way of integrating ecological processes and evolutionary processes.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
